---
title: "Generating a static STAC catalog with rstatic"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generating a static STAC catalog with rstatic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# Some chunks use the optional 'terra' package.
has_terra <- requireNamespace("terra", quietly = TRUE)
```

```{r setup}
library(rstatic)
```

## Overview

`rstatic` provides composable primitives to build and serialize **static**
SpatioTemporal Asset Catalog (STAC) documents. A static catalog is a tree of
plain JSON files linked to one another, following the
[STAC specification](https://github.com/radiantearth/stac-spec) version
1.0.0:

```
stac/
  catalog.json
  collections/
    {collection}/
      collection.json
      items/
        {item}/
          item.json
```

This vignette shows how to build such a tree from scratch. The design
separates a **pure core** that builds documents in memory from a thin
**I/O shell**: [stac_save()] is the writer and [stac_read()] the
reader. Nothing is written until you call `stac_save()` at the end.

To be fully self-contained in this vignette, all output is written under a temporary directory:

```{r root}
root <- tempfile("stac-")
```

## 1. Build the root catalog

`new_catalog()` builds the root catalog document.

```{r catalog}
catalog <- new_catalog(
  id = "rstatic-catalog",
  title = "rstatic package catalog",
  description = "An example static STAC catalog built with rstatic"
)
catalog
```

## 2. Create and attach a collection

In STAC, collections group related items. `new_collection()` builds the
document and `add_collection()` registers it as a `child` of the catalog,
returning the updated catalog. No files are written. Any extra named 
arguments (e.g. `citation`, `doi`) are stored as additional collection
fields.

```{r collection}
collection <- new_collection(
  id = "land-cover",
  title = "Example Land Cover",
  description = "Annual land cover map (example data)",
  license = "CC-BY-4.0"
)
catalog <- add_collection(catalog, collection)
collection
```

## 3. Spatial metadata

If [terra](https://rspatial.github.io/terra/) is installed, `extract_bbox()`
reads a raster and returns its bounding box in WGS84. When `terra` is not
available, you can pass a bounding box directly.

A small Sentinel-2 subset ships with the package for demonstration.

```{r bbox, eval = has_terra}
tif_path <- system.file(
  "extdata/lulc/S2_MSI_20LMR_2022-01-15_2022-12-23_class_v1.tif",
  package = "rstatic"
)
bbox <- extract_bbox(tif_path)
bbox
```

The `as_geometry()` helper turns a bounding box into a `GeoJSON polygon` and
has no external dependencies:

```{r geometry}
bbox <- c(-63.63658, -8.630256, -63.41822, -8.412882)
geometry <- as_geometry(bbox)
geometry
```

## 4. Build items

Items are `GeoJSON Feature`s. Build their `properties` with
`new_properties()` and `assets` with `new_asset()`; the asset media type is
deduced from the file extension.

```{r item}
item <- new_item(
  id = "land-cover-2022",
  bbox = bbox,
  geometry = geometry,
  properties = new_properties(
    datetime = "2022-07-16T00:00:00Z",
    description = "Example land cover map for 2022"
  ),
  assets = list(
    data = new_asset(tif_path, title = "Land Cover 2022")
  )
)
item
```

## 5. Add items to the collection

`add_items()` links each item from the collection and updates the collection's
spatial and temporal extent automatically. It returns the updated collection
without writing anything on disk.

```{r add-items}
collection <- add_items(collection, item)
print(collection$extent$spatial$bbox[[1]])
print(collection$extent$temporal$interval[[1]])
```

## 6. Thumbnails and styles (optional)

`new_thumbnail()` describes a `PNG` preview to render from a raster. It is 
a pure builder: it reads no raster and writes no file. It returns a STAC 
asset (like `new_asset()`) carrying the render *intent* (i.e. the source raster, width, and style) which `stac_save()` materializes later, when the
owning item is written. A `stac_style()` object describes how raster values 
map to thumbnail pixels; it only validates and normalizes the rendering intent.

The LULC raster shipped with the package
(`S2_MSI_20LMR_2022-01-05_2022-12-23_class_v1.tif`) is a single band `GeoTIFF`
classification map for 2022 based on Sentinel-2 L2A data cube, tile `20LMR` 
in Rondônia state, Brazil. The data cube and training points are available in
[`sitsdata` repository](https://github.com/e-sensing/sitsdata).

A categorical style type is inferred from the parameters.
Supplying `values` and `colors` produces a categorical style, suitable
for a land-cover map where each integer class has its own color:

```{r style-categorical}
land_cover_style <- stac_style(
  values = c(1, 2, 3, 4, 5, 6, 7, 8, 9),
  colors = c("#F39C12", "#CD6155", "#E0DB34", "#1E8449", "#229C59", 
             "#00B29E", "#3ABABA", "#2980B9", "#A0B9C8"),
  labels = c("Clear_Cut_Bare_Soil", "Clear_Cut_Burned_Area", 
             "Clear_Cut_Vegetation", "Forest", "Mountainside_Forest", 
             "Riparian_Forest", "Seasonally_Flooded", "Water", "Wetland"),
  nodata = 0
)
land_cover_style
```

Attach the thumbnail intent to the item under the `"thumbnail"` key, then add
the updated item to the collection. No raster is touched yet. The `PNG` is
rendered at save time, which requires `terra`, so we only attach it when
`terra` is available:

```{r thumbnail, eval = has_terra}
thumb <- new_thumbnail(
  item$assets$data,
  width = 200,
  style = land_cover_style
)
item <- add_asset(item, "thumbnail", thumb)
collection <- add_items(collection, item)
```

`qml_to_style()` reads a supported QGIS `.qml` raster style and returns the
same kind of object as `stac_style()` (it requires `xml2`); a QML-styled
thumbnail is attached in exactly the same way.

```{r qml-style, eval = requireNamespace("xml2", quietly = TRUE)}
qml_style <- qml_to_style(
  system.file("extdata/lulc/S2_MSI_20LMR_2022-01-05_2022-12-23_class_v1.qml",
              package = "rstatic")
)
qml_style
```

## 7. Save the catalog

So far nothing has been written. `stac_save()` is the only writer: it persists
exactly the documents you hand it. It proceeds a pure overwrite, with no
implicit reads or merges, and renders any thumbnail intent into the item's
directory. Documents are written children-first (items, then collection, then
catalog).

```{r save}
stac_save(catalog = catalog, collection = collection, items = item,
          root_dir = root)
```

When a thumbnail was attached, its PNG now exists under the item directory and
can be plotted:

```{r thumbnail-plot, eval = has_terra && requireNamespace("png", quietly = TRUE), fig.cap = "Grayscale thumbnail of the Sentinel-2 B04 band."}

wd <- getwd()
setwd(root)
plot(item$assets$thumbnail)
setwd(wd)
```

## 8. Update a catalog already on disk

To add to a catalog that is already persisted -- typically from a *separate*
script populating the same catalog over time -- read it back, build in memory,
and save again. Because `stac_save()` never merges implicitly, reading first is
what preserves the children already registered on disk:

```{r accumulate}
catalog <- stac_read(
  file.path(root, "stac", "catalog.json"),
  default = new_catalog("restore-plus", "Restore+ Catalog", "...")
)

deforestation <- new_collection(
  id = "deforestation",
  title = "Example Deforestation",
  description = "A second collection added in a later run"
)
catalog <- add_collection(catalog, deforestation)
stac_save(catalog = catalog, collection = deforestation, root_dir = root)

# The catalog now links both collections.
Filter(function(l) l$rel == "child", catalog$links)
```

## Resulting catalog

The final directory tree contains the linked JSON documents:

```{r tree}
list.files(file.path(root, "stac"), recursive = TRUE)
```

Each file is a self-contained STAC document. For example, the item:

```{r item-json}
cat(readLines(
  file.path(root, "stac", "collections", "land-cover",
            "items", "land-cover-2020", "item.json")
), sep = "\n")
```
