---
title: "Generating a static STAC catalog with rstatic"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generating a static STAC catalog with rstatic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# Some chunks use the optional 'terra' package.
has_terra <- requireNamespace("terra", quietly = TRUE)
```

```{r setup}
library(rstatic)
```

## Overview

`rstatic` provides composable primitives to build and serialize **static**
SpatioTemporal Asset Catalog (STAC) documents. A static catalog is a tree of
plain JSON files linked to one another, following the
[STAC specification](https://github.com/radiantearth/stac-spec) version
1.0.0:

```
stac/
  catalog.json
  collections/
    {collection}/
      collection.json
      items/
        {item}/
          item.json
```

This vignette walks through building such a tree from scratch. All output is
written under a temporary directory so the vignette is fully self-contained.

```{r root}
root <- tempfile("stac-")
```

## 1. Initialize the root catalog

`stac_init()` creates (or updates) the root `catalog.json`. It preserves any
existing `child` links, so it is safe to call repeatedly.

```{r catalog}
catalog <- stac_init(
  id = "restore-plus",
  title = "Restore+ Catalog",
  description = "An example static STAC catalog built with rstatic",
  root_dir = root
)
catalog
```

## 2. Create and register a collection

Collections group related items. `new_collection()` builds the document in
memory and `stac_add_collection()` persists it, links it to the catalog, and
returns the updated catalog. Any extra named arguments (e.g. `citation`, `doi`)
are stored as additional collection fields.

```{r collection}
collection <- new_collection(
  id = "land-cover",
  title = "Example Land Cover",
  description = "Annual land cover maps (example data)",
  license = "CC-BY-4.0"
)
catalog <- stac_add_collection(catalog, collection = collection, root_dir = root)
collection
```

## 3. Spatial metadata

If [terra](https://rspatial.github.io/terra/) is installed, `extract_bbox()`
reads a raster and returns its bounding box in WGS84. A small Sentinel-2
subset ships with the package for demonstration.

```{r bbox, eval = has_terra}
tif <- system.file("extdata/example.tif", package = "rstatic")
bbox <- extract_bbox(tif)
bbox
```

When `terra` is not available, you can pass a bounding box directly. The
`as_geometry()` helper turns a bounding box into a GeoJSON polygon and has no
external dependencies:

```{r geometry}
bbox <- c(-50, -10, -49, -9)
geometry <- as_geometry(bbox)
str(geometry, max.level = 2)
```

## 4. Build items

Items are GeoJSON `Feature`s. Build their `properties` with
`new_properties()` and `assets` with `new_asset()`; the asset media type is
deduced from the file extension.

```{r item}
item <- new_item(
  id = "land-cover-2020",
  bbox = bbox,
  geometry = geometry,
  properties = new_properties(
    datetime = "2020-01-01T00:00:00Z",
    description = "Example land cover map for 2020"
  ),
  assets = list(
    data = new_asset("land-cover-2020.tif", title = "Land Cover 2020")
  )
)
item
```

## 5. Add items to the collection

`stac_add_items()` writes each item, links it from the collection, and updates
the collection's spatial and temporal extent automatically.

```{r add-items}
collection <- stac_add_items(collection, item, root_dir = root)
collection$extent$spatial$bbox[[1]]
collection$extent$temporal$interval[[1]]
```

## 6. Thumbnails and styles (optional)

With `terra`, `new_thumbnail()` renders a PNG preview. A `stac_style()` object
describes how raster values map to thumbnail pixels: it does not render the
image, it only validates and normalizes the rendering intent.

The raster shipped with the package (`example.tif`) is a single Landsat
surface-reflectance band (`B04`) with values roughly in the `[0, 0.5]` range.
A continuous grayscale style stretches those values between black and white:

```{r style}
style <- stac_style(min = 0, max = 0.5, palette = c("black", "white"))
style
```

The style kind is inferred from the parameters rather than chosen explicitly.
Supplying `values` and `colors` instead produces a categorical style, suitable
for a land-cover map where each integer class has its own color:

```{r style-categorical}
land_cover_style <- stac_style(
  values = c(1, 2, 3, 100, 101, 102),
  colors = c("#c14d00", "#367906", "#7cc900", "#f00060", "#4cdcc7", "#cbba6b"),
  labels = c("Annual crop", "Cerradão", "Cerrado", "Urban",
             "Silviculture", "Natural non-vegetated"),
  nodata = 0
)
land_cover_style
```

`new_thumbnail()` applies the style with `terra` to write the PNG. Here we
render the Landsat band with the grayscale style:

```{r thumbnail, eval = has_terra, fig.cap = "Grayscale thumbnail of the Landsat B04 band."}
thumb <- new_thumbnail(
  collection_id = "land-cover",
  item_id = "land-cover-2020",
  asset_href = system.file("extdata/example.tif", package = "rstatic"),
  width = 200,
  style = style,
  root_dir = root
)
plot(thumb)
```

`qml_to_style()` reads a supported QGIS `.qml` raster style and returns the
same kind of object as `stac_style()` (it requires `xml2`). The bundled
`example.qml` is a single-band pseudocolor style, which becomes a continuous
style with a color ramp:

```{r qml-style, eval = requireNamespace("xml2", quietly = TRUE)}
qml_style <- qml_to_style(system.file("extdata/example.qml", package = "rstatic"))
qml_style
```

```{r thumbnail-qml, eval = has_terra && requireNamespace("xml2", quietly = TRUE), fig.cap = "Thumbnail rendered from a QML style."}
thumb_qml <- new_thumbnail(
  collection_id = "land-cover",
  item_id = "land-cover-2020-qml",
  asset_href = system.file("extdata/example.tif", package = "rstatic"),
  width = 200,
  style = qml_style,
  root_dir = root
)
plot(thumb_qml)
```

## Resulting catalog

The final directory tree contains the linked JSON documents:

```{r tree}
list.files(file.path(root, "stac"), recursive = TRUE)
```
