---
title: "Generating a static STAC catalog with rstatic"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generating a static STAC catalog with rstatic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# Some chunks use the optional 'terra' package.
has_terra <- requireNamespace("terra", quietly = TRUE)
```

```{r setup}
library(rstatic)
```

## Overview

`rstatic` provides composable primitives to build and serialize **static**
SpatioTemporal Asset Catalog (STAC) documents. A static catalog is a tree of
plain JSON files linked to one another, following the
[STAC specification](https://github.com/radiantearth/stac-spec) version
1.0.0:

```
stac/
  catalog.json
  collections/
    {collection}/
      collection.json
      items/
        {item}/
          item.json
```

This vignette walks through building such a tree from scratch. All output is
written under a temporary directory so the vignette is fully self-contained.

```{r root}
root <- tempfile("stac-")
```

## 1. Initialize the root catalog

`stac_init()` creates (or updates) the root `catalog.json`. It preserves any
existing `child` links, so it is safe to call repeatedly.

```{r catalog}
catalog <- stac_init(
  id = "restore-plus",
  title = "Restore+ Catalog",
  description = "An example static STAC catalog built with rstatic",
  root_dir = root
)
catalog
```

## 2. Create and register a collection

Collections group related items. `new_collection()` builds the document in
memory and `stac_add_collection()` persists it, links it to the catalog, and
returns the updated catalog. Any extra named arguments (e.g. `citation`, `doi`)
are stored as additional collection fields.

```{r collection}
collection <- new_collection(
  id = "land-cover",
  title = "Example Land Cover",
  description = "Annual land cover maps (example data)",
  license = "CC-BY-4.0"
)
catalog <- stac_add_collection(catalog, collection = collection, root_dir = root)
collection
```

## 3. Spatial metadata

If [terra](https://rspatial.github.io/terra/) is installed, `extract_bbox()`
reads a raster and returns its bounding box in WGS84. A small Sentinel-2
subset ships with the package for demonstration.

```{r bbox, eval = has_terra}
tif <- system.file("extdata/example.tif", package = "rstatic")
bbox <- extract_bbox(tif)
bbox
```

When `terra` is not available, you can pass a bounding box directly. The
`as_geometry()` helper turns a bounding box into a GeoJSON polygon and has no
external dependencies:

```{r geometry}
bbox <- c(-50, -10, -49, -9)
geometry <- as_geometry(bbox)
str(geometry, max.level = 2)
```

## 4. Build items

Items are GeoJSON `Feature`s. Build their `properties` with
`new_properties()` and `assets` with `new_asset()`; the asset media type is
deduced from the file extension.

```{r item}
item <- new_item(
  id = "land-cover-2020",
  bbox = bbox,
  geometry = geometry,
  properties = new_properties(
    datetime = "2020-01-01T00:00:00Z",
    description = "Example land cover map for 2020"
  ),
  assets = list(
    data = new_asset("land-cover-2020.tif", title = "Land Cover 2020")
  )
)
item
```

## 5. Add items to the collection

`stac_add_items()` writes each item, links it from the collection, and updates
the collection's spatial and temporal extent automatically.

```{r add-items}
collection <- stac_add_items(collection, item, root_dir = root)
collection$extent$spatial$bbox[[1]]
collection$extent$temporal$interval[[1]]
```

## 6. Thumbnails and styles (optional)

With `terra`, `new_thumbnail()` renders a PNG preview. Styles can be built
explicitly with `stac_style()` or parsed from a QGIS `.qml` file with
`qml_to_style()` (requires `xml2`).

```{r style}
style <- stac_style(min = 0, max = 255, palette = c("black", "white"))
style$type
```

```{r thumbnail, eval = has_terra, fig.cap = "Rendered thumbnail for the land-cover-2020 item."}
thumb <- new_thumbnail(
  collection_id = "land-cover",
  item_id = "land-cover-2020",
  asset_href = system.file("extdata/example.tif", package = "rstatic"),
  width = 200,
  style = style,
  root_dir = root
)
plot(thumb)
```

## Resulting catalog

The final directory tree contains the linked JSON documents:

```{r tree}
list.files(file.path(root, "stac"), recursive = TRUE)
```
